---
title: "Tessera + Seurat: Spatial Clustering Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tessera + Seurat: Spatial Clustering Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width='80%',
  fig.align='center'
)
```

# Overview

This vignette shows how to use **Tessera** tiles as the input to a standard
**Seurat V5** clustering workflow.  The key idea is that Tessera partitions a
spatial transcriptomics sample into contiguous tiles of 5–50 cells, and each
tile can then be treated as one observation in Seurat — analogous to a single
cell in scRNA-seq.  This unlocks all of Seurat's downstream tools (PCA, UMAP,
`FindClusters`, differential expression, etc.) at the tile level.

**Workflow at a glance:**

1. Run Tessera on raw coordinates + counts → tiles
2. Create a Seurat object from tile-level counts
3. Cluster tiles in embedding space
4. Visualise clusters in UMAP and physical space
5. Transfer tile labels back to individual cells
6. Inspect spatial cluster composition and gene expression

# Libraries

```{r}
suppressPackageStartupMessages({
    library(tessera)

    ## Downstream analysis
    library(Seurat)

    ## Plotting (not imported by tessera)
    library(ggplot2)
    library(ggthemes)
    library(viridis)
    library(patchwork)
})

## Jupyter plot sizing helper (uses repr package under the hood)
fig.size <- function(h, w) {
    options(repr.plot.height = h, repr.plot.width = w)
}

## Helper: convert sf tile shapes to geom_polygon-compatible data.table.
## We use geom_polygon + coord_fixed instead of geom_sf + coord_sf because
## the latter requires GEOS to compute graticule lengths, which can fail on
## planar (non-geographic) coordinates.
sf_to_poly <- function(df, shape_col = "shape") {
    shapes <- df[[shape_col]]
    coords_list <- lapply(seq_along(shapes), function(i) {
        co <- sf::st_coordinates(shapes[[i]])
        data.table::data.table(x = co[, 1], y = co[, 2], .tile_group = i)
    })
    coords_dt <- data.table::rbindlist(coords_list)
    meta <- data.table::as.data.table(df)[, !..shape_col]
    meta[, .tile_group := .I]
    merge(coords_dt, meta, by = ".tile_group")
}

```

---

# Load data

Small MERFISH dataset (3 177 cells, 479 genes) from
[Chen et al. (2023)](https://www.biorxiv.org/content/10.1101/2023.04.04.535379v1.abstract).

```{r}
data('tessera_warmup')
counts    = tessera_warmup$counts
meta_data = tessera_warmup$meta_data
```

Each cell has spatial coordinates (`X`, `Y`) and a coarse cell-type label
(`type`) that we will use later to interpret tile clusters.

```{r}
dim(counts)           # genes x cells
head(meta_data)
table(meta_data$type)
```

```{r fig.height = 8, fig.width = 8}
fig.size(8, 8)
ggplot() +
    geom_point(data = meta_data, aes(X, Y, color = type)) +
    theme_void() +
    scale_color_tableau() +
    coord_fixed() +
    labs(title = 'Cell types in physical space') +
    NULL
```

---

# Step 1 — Run Tessera

`RunTessera()` is the one-call wrapper that runs the full pipeline:
`make_cells → make_mesh → compute_field → compute_morse → run_dmt → make_tiles`.

It returns two objects:

| Object | Description |
|--------|-------------|
| `dmt` | Cell-level data: coordinates, mesh, PCA, and tile assignment (`dmt$pts$agg_id`) |
| `aggs` | Tile-level data: pooled counts, PCA embeddings, shapes, adjacency |

`meta_vars_include` specifies which columns of `meta_data` to carry forward
into the tile-level summaries (here, cell-type proportions per tile).

```{r}
res = RunTessera(
    X = meta_data$X,
    Y = meta_data$Y,
    counts = counts,
    meta_data = meta_data,
    meta_vars_include = c('type'),
)
dmt  = res$dmt
aggs = res$aggs
```

The first three tile PCs capture the dominant spatial patterns of gene
expression variation.  Each tile is coloured by its PC score:

```{r fig.height = 10, fig.width = 30}
fig.size(10, 30)
purrr::map(1:3, function(i) {
    ggplot(sf_to_poly(cbind(aggs$meta_data, val=aggs$pcs[, i]))) +
        geom_polygon(aes(x, y, group = .tile_group, fill = val)) +
        theme_void(base_size = 16) +
        coord_fixed() +
        scale_fill_gradient2_tableau() +
        guides(color = 'none') +
        labs(title = paste0('PC', i)) +
        NULL
}) %>%
    purrr::reduce(`|`)
```

---

# Step 2 — Build a Seurat object from tiles

We treat each tile as an observation, just like a cell in scRNA-seq.

* **Counts:** `aggs$counts` — genes × tiles sparse matrix of pooled raw counts.
* **Metadata:** tile-level summaries (centroid, `npts`, area, etc.).
* **PCA embeddings:** `aggs$pcs` — each tile represented by the mean of its
  constituent cells' PCA scores.  This tends to give more biologically
  meaningful clusters than re-running PCA on pooled counts, because it
  preserves the continuous cell-level variation.

```{r}
## Create Seurat object from tile counts
obj = Seurat::CreateSeuratObject(
    counts = aggs$counts,
    meta.data = tibble::column_to_rownames(
        data.frame(dplyr::select(aggs$meta_data, -shape)),
        'id'
    )
)

## Store sf shapes separately (Seurat metadata doesn't handle sf well)
obj@meta.data$shape = aggs$meta_data$shape

## Inject pre-computed PCA as a DimReducObject
rownames(aggs$pcs) = colnames(obj)
obj[['pca']] = Seurat::CreateDimReducObject(
    embeddings = aggs$pcs,
    loadings   = dmt$udv_cells$loadings,
    key        = 'pca_',
    assay      = Seurat::DefaultAssay(obj)
)
```

---

# Step 3 — Cluster tiles with Seurat

Standard Seurat pipeline: normalise counts → UMAP → shared-nearest-neighbour
graph → Leiden/Louvain clustering.

```{r}
.verbose = FALSE
obj = obj %>%
    NormalizeData(
        normalization.method = 'LogNormalize',
        scale.factor = median(obj@meta.data$nCount_RNA),
        verbose = .verbose
    ) %>%
    RunUMAP(verbose = .verbose, dims = 1:10, reduction = 'pca') %>%
    Seurat::FindNeighbors(features = 1:10, reduction = 'pca', verbose = .verbose) %>%
    Seurat::FindClusters(verbose = .verbose, resolution = c(2))
```

```{r}
cat('Number of tile clusters:', length(levels(obj@meta.data$seurat_clusters)), '\n')
table(obj@meta.data$seurat_clusters)
```

---

# Step 4 — Visualise tile clusters

Side-by-side: UMAP embedding (left) and physical tissue space (right).

```{r fig.height = 6, fig.width = 12}
p1 = DimPlot(obj, reduction = 'umap', group.by = 'seurat_clusters') +
    scale_color_tableau('Classic 10')
p2 = ggplot(sf_to_poly(obj@meta.data)) +
    geom_polygon(aes(x, y, group = .tile_group, fill = seurat_clusters)) +
    theme_void(base_size = 16) +
    coord_fixed() +
    scale_fill_tableau('Classic 10') +
    labs(title = 'Tile clusters in tissue space') +
    NULL

fig.size(6, 12)
(p1 | p2) + plot_layout(widths = c(1, 1))
```

---

# Step 5 — Transfer tile labels to cells

Every surviving cell in `dmt$pts` has an `agg_id` pointing to its tile.  We
use this to propagate tile-level cluster labels down to individual cells.

```{r}
dmt$pts$spatial_cluster = obj@meta.data$seurat_clusters[dmt$pts$agg_id]
```

Now we can overlay cell-type annotations with tile boundaries and cluster
colours to see how well the spatial clusters capture known biology.

```{r fig.height = 8, fig.width = 20}
tile_poly = sf_to_poly(obj@meta.data)
p1 = ggplot() +
    geom_polygon(data = tile_poly, aes(x, y, group = .tile_group),
                 fill = NA, color = 'grey70') +
    geom_point(data = dmt$pts, aes(X, Y, color = type)) +
    scale_color_tableau() +
    theme_void() +
    coord_fixed() +
    labs(title = 'Cell types + tile boundaries') +
    NULL
p2 = ggplot() +
    geom_polygon(data = tile_poly,
                 aes(x, y, group = .tile_group, fill = seurat_clusters),
                 alpha = .2) +
    geom_point(data = dmt$pts, aes(X, Y, color = spatial_cluster)) +
    scale_color_tableau('Classic 10') +
    scale_fill_tableau('Classic 10') +
    theme_void() +
    guides(fill = 'none') +
    coord_fixed() +
    labs(title = 'Spatial clusters (tiles + cells)') +
    NULL
fig.size(8, 20)
p1 | p2
```

---

# Step 6 — Interpret spatial clusters

## Cell-type composition

A stacked bar plot showing what fraction of each spatial cluster is made up
of each cell type.  This tells us whether the spatial clusters are driven by
cell-type composition (e.g., immune-enriched vs. epithelial regions).

```{r fig.height = 8, fig.width = 10}
fig.size(8, 10)
dmt$pts %>%
    with(table(type, spatial_cluster)) %>%
    prop.table(2) %>%
    data.table() %>%
    ggplot(aes(spatial_cluster, 100 * N, fill = type)) +
        geom_bar(stat = 'identity', position = position_stack()) +
        scale_fill_tableau() +
        theme_bw(base_size = 20) +
        labs(y = '% of spatial cluster', fill = 'cell type',
             title = 'Cell-type composition per spatial cluster') +
        NULL
```

## Tile-level gene expression

We can query individual genes at the tile level, just as we would at the
cell level in standard scRNA-seq.  Each tile is coloured by its normalised
expression of the queried gene.

```{r fig.height = 8, fig.width = 10}
feature = 'MKI67'    ## marker for dividing cells

fig.size(8, 10)
ggplot(sf_to_poly(cbind(obj@meta.data, FetchData(obj, feature)))) +
    geom_polygon(aes(x, y, group = .tile_group, fill = !!sym(feature))) +
    scale_fill_gradient(low = 'white', high = '#832424') +
    theme_void() +
    coord_fixed() +
    labs(title = paste0('Tile-level expression: ', feature)) +
    NULL

```

---

# Session Info

```{r}
sessionInfo()
```
