---
title: "Walkthrough: Tessera Algorithm Step-by-Step"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Walkthrough: Tessera Algorithm Step-by-Step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width='80%',
  fig.align='center'
)
```

# Overview

This notebook is a walkthrough of the steps of Tessera, for those who want to get more familiar with the components.


# Libs

```{r}
suppressPackageStartupMessages({
    library(tessera)

    ## Plotting functions
    ## Not imported by Tessera
    library(ggplot2)
    library(ggthemes)
    library(viridis)
    library(patchwork)
})

fig.size <- function(h, w) {
    options(repr.plot.height = h, repr.plot.width = w)
}

```

# Parameters

```{r}
verbose = TRUE
show_plots = TRUE

###### STEP 0 ######
npcs = 10
## Graph pruning
prune_thresh_quantile = 0.95
prune_min_cells = 10


###### STEP 1: GRADIENTS ######
smooth_distance = c('none', 'euclidean', 'projected', 'constant')[3]
smooth_similarity = c('none', 'euclidean', 'projected', 'constant')[3]


###### STEP 2: DMT ######
## ... no options


###### STEP 3: AGGREGATION ######
max_npts = 50
min_npts = 5
alpha = 1 ## 0.2 = conservative merging, 2 = liberal merging

```

# Data

Small sample MERFISH dataset from Chen et al:

https://www.biorxiv.org/content/10.1101/2023.04.04.535379v1.abstract

```{r}
data('tessera_warmup')
counts = tessera_warmup$counts
meta_data = tessera_warmup$meta_data
meta_vars_include = c('type')
```

```{r fig.height = 8, fig.width = 8}
fig.size(8, 8)
ggplot() +
    geom_point(data = meta_data, aes(X, Y, color = type)) +
    theme_void() +
    scale_color_tableau() +
    coord_fixed() +
    NULL
```

Some coarse and fine grained cell types are predefined here, to help interpret the tiles we get below.

```{r}
table(meta_data$type)
```

## prepare

```{r}
coords = matrix(c(meta_data$X, meta_data$Y), ncol = 2, dimnames = list(NULL, c('x', 'y')))
cells = make_cells(coords, NULL, counts, data.table::as.data.table(meta_data))
mesh = make_mesh(cells, list(
    prune_thresh_quantile = prune_thresh_quantile,
    prune_mincells        = prune_min_cells
))
```

```{r fig.height = 15, fig.width = 15}
fig.size(15, 15)
# fig.size(20, 20)

if (show_plots) {
    ggplot() +

        # ## big data
        # geom_point(data = mesh$pts, aes(X, Y), shape = '.', alpha = .1) +
        # geom_segment(data = mesh$edges[boundary == TRUE], aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt), color = 'red', lwd = .2) +

        ## small data
        geom_segment(data = mesh$edges, aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt), color = 'black', lwd = .2) +
        geom_segment(data = mesh$edges[boundary == TRUE], aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt), color = 'red', lwd = .2) +
        geom_point(data = mesh$pts, aes(X, Y), size = .5) +

        theme_void(base_size = 20) +
        coord_cartesian(expand = FALSE) +
        labs(title = 'Pruned adjacency graph') +
        NULL
}
```

## pca

```{r}
pca = do_pca(cells$counts[, mesh$pts$ORIG_ID, drop = FALSE], npcs)
cells$embeddings = list(pca = pca$embeddings)
```


# Step 1: compute gradients on all data structures

```{r}
field = compute_field(cells, mesh, list(
    smooth_distance   = smooth_distance,
    smooth_similarity = smooth_similarity
))
```

```{r fig.height = 15, fig.width = 15}
if (show_plots) {

    len_plot_constant = .8
    fig.size(15, 15)
    # fig.size(20, 20)
    ggplot() +
        geom_point(data = mesh$pts, aes(X, Y), size = .5) +
        geom_segment(data = mesh$edges[boundary == TRUE, ], aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt), color = 'red') +

        ## Triangle Gradients
        geom_segment(
            data = data.table(mesh$triangles, field$gradient$tris),
            aes(
                x=X-len_plot_constant*(len_grad+len_ortho)*dx_ortho,
                y=Y-len_plot_constant*(len_grad+len_ortho)*dy_ortho,
                xend=X+len_plot_constant*(len_grad+len_ortho)*dx_ortho,
                yend=Y+len_plot_constant*(len_grad+len_ortho)*dy_ortho
            ),
            linewidth = .4, alpha = 1,
            color = 'blue'
        ) +

        theme_void() +
        coord_fixed(expand = FALSE) +
        NULL
}
```

# Step 2: DMT

```{r}
mesh$morse = compute_morse(field, mesh)
```

## f values

```{r fig.height = 15, fig.width = 15}
if (show_plots) {
    ntri = max(which(mesh$triangles$external == FALSE))
    i = Matrix::t(mesh$tri_to_pt[1:ntri, ])@i+1
    plt_df = data.table(
        X = mesh$pts$X[i],
        Y = mesh$pts$Y[i],
        f = rep(mesh$morse$f$tris[1:ntri], each = 3)
    )[
        , id := rep(1:ntri, each = 3)
    ][]


    fig.size(15, 15)
    ggplot() +
        geom_polygon(data = plt_df, aes(X, Y, group = id, fill = f, color = f)) +
        theme_void() +
        coord_fixed(expand = FALSE) +
        scale_fill_viridis() +
        scale_color_viridis() +
        NULL
}
```

## forests

```{r fig.height = 15, fig.width = 15}
if (show_plots) {
    fig.size(15, 15)
    ggplot() +
        ## dual forest
        geom_point(data = mesh$triangles[mesh$morse$dual$maxima, ], aes(X, Y), color = 'blue', size = 2) +
        geom_segment(data = mesh$morse$dual$edges, aes(x=x0, y=y0, xend=x1, yend=y1), color = 'blue') +

        ## primary forest
        geom_point(data = mesh$pts[mesh$morse$prim$minima, ], aes(X, Y), color = 'red', size = 2) +
        geom_segment(data = mesh$morse$prim$edges, aes(x=x0, y=y0, xend=x1, yend=y1), color = 'red') +

        theme_void() +
        coord_cartesian(expand = FALSE) +
        NULL
}
```

## separatrices

After DMT, we have a separatrix (in blue) and the primary forest connecting points (in red).

```{r fig.height = 15, fig.width = 15}
if (show_plots) {
    fig.size(15, 15)
    ggplot() +

        geom_segment(data = mesh$edges[mesh$morse$e_sep, ], aes(x = x0_tri, y = y0_tri, xend = x1_tri, yend = y1_tri), lwd = 1, color = 'blue') +
        geom_segment(data = mesh$edges[boundary == TRUE], aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt), color = 'blue', lwd = 1) +

        ## primary forest
        geom_point(data = mesh$pts[mesh$morse$prim$minima, ], aes(X, Y), color = 'red', size = 2) +
        geom_segment(data = mesh$morse$prim$edges, aes(x=x0, y=y0, xend=x1, yend=y1), color = 'red') +

        theme_void() +
        coord_cartesian(expand = FALSE) +
        NULL
}
```

## extract tiles

```{r}
dmt_run = run_dmt(mesh)
```

```{r fig.height = 15, fig.width = 20}
if (show_plots) {
    set.seed(2)
    fig.size(15, 20)
    ggplot() +
        geom_point(data = dmt_run$pts, aes(X, Y, color = factor(agg_id)), size = 1) +
        theme_void() +
        coord_cartesian(expand = FALSE) +
        guides(color = 'none') +
        NULL
}
```

```{r fig.height = 15, fig.width = 20}
if (show_plots) {
    pts_with_meta = data.table::copy(dmt_run$pts)
    pts_with_meta[, type := meta_data$type[ORIG_ID]]
    set.seed(2)
    fig.size(15, 20)
    ggplot() +
        geom_point(data = pts_with_meta, aes(X, Y, color = type)) +
        theme_void() +
        coord_cartesian(expand = FALSE) +
        scale_color_tableau() +
        guides(color = 'none') +
        NULL
}
```

# Step 3: Aggregation

Merge similar aggregates that are nearby, then clean up small tiles below `min_npts`.

```{r}
tiles = make_tiles(cells, mesh, dmt_run, list(
    alpha    = alpha,
    max_npts = max_npts,
    min_npts = min_npts
))
```

## Final tiles

```{r fig.height = 10, fig.width = 30}
fig.size(10, 30)
if (show_plots) {
    purrr::map(1:3, function(i) {
        ggplot(cbind(tiles$meta_data, val=tiles$pcs[, i])) +
            geom_sf(aes(geometry = shape, fill = val)) +
            theme_void(base_size = 16) +
            coord_sf(datum = NA) +
            scale_fill_gradient2_tableau() +
            guides(color = 'none') +
            labs(title = paste0('PC', i)) +
            NULL
    }) %>%
        purrr::reduce(`|`)
}
```

# Results

## Tiles

The primary output is the tiles. Each tile has a row in the meta_data table:

- npts denotes the number of cells in the tile.


```{r}
head(tiles$meta_data)
```

We also have pooled gene counts, for differential gene expression analysis.

```{r}
tiles$counts[1:5, 1:5]
```

And we have PCA embeddings for the tiles.

```{r}
head(tiles$pcs)
```

The rest of the fields are internal to the algorithm and can be ignored.

```{r}
setdiff(names(tiles), c('pcs', 'meta_data', 'counts'))
```

## Points

We also keep information about the cells in `tiles$cell_ids`.

```{r}
names(tiles)
```

The important fields to keep for further analyses are:

- ORIG_ID: the index of the cell in the input data. Some cells get filtered out as outliers, so not all input cells get assigned to a tile.
- tile_id: the index of the tile in the `tiles` data structures above.


```{r}
head(tiles$cell_ids)
```

# Session Info

```{r}
sessionInfo()
```

