---
title: "Walkthrough: Tessera Algorithm Step-by-Step"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Walkthrough: Tessera Algorithm Step-by-Step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width='80%',
  fig.align='center'
)
```

# Overview

This notebook is a walkthrough of the steps of Tessera, for those who want to get more familiar with the components. 


# Libs

```{r}
suppressPackageStartupMessages({
    library(tessera)

    ## Plotting functions 
    ## Not imported by Tessera
    library(ggplot2)
    library(ggthemes)
    library(viridis)
    library(patchwork)
})

fig.size <- function(h, w) {
    options(repr.plot.height = h, repr.plot.width = w)
}

```

# Parameters

```{r}
verbose = TRUE
show_plots = TRUE

###### STEP 0 ######
npcs = 10
## Graph pruning
prune_thresh_quantile = 0.95
prune_min_cells = 10


###### STEP 1: GRADIENTS ######
smooth_distance = c('none', 'euclidean', 'projected', 'constant')[3] 
smooth_similarity = c('none', 'euclidean', 'projected', 'constant')[3] 


###### STEP 2: DMT ######
## ... no options


###### STEP 3: AGGREGATION ######
max_npts = 50
min_npts = 5
alpha = 1 ## 0.2 = conservative merging, 2 = liberal merging 

```

# Data

Small sample MERFISH dataset from Chen et al: 

https://www.biorxiv.org/content/10.1101/2023.04.04.535379v1.abstract

```{r}
data('tessera_warmup')
counts = tessera_warmup$counts
meta_data = tessera_warmup$meta_data
meta_vars_include = c('type')
```

```{r fig.height = 8, fig.width = 8}
fig.size(8, 8)
ggplot() + 
    geom_point(data = meta_data, aes(X, Y, color = type)) + 
    theme_void() + 
    scale_color_tableau() + 
    coord_sf(expand = FALSE) + 
    NULL
```

Some coarse and fine grained cell types are predefined here, to help interpret the tiles we get below. 

```{r}
table(meta_data$type)
```

## prepare

```{r}
dmt = init_data(meta_data$X, meta_data$Y, counts, meta_data, meta_vars_include)
dmt = prune_graph(dmt, thresh_quantile = prune_thresh_quantile, mincells = prune_min_cells) 
```

```{r}
dmt = add_exterior_triangles(dmt)
```

```{r fig.height = 15, fig.width = 15}
fig.size(15, 15)
# fig.size(20, 20)

if (show_plots) {    
    ggplot() + 
    
        # ## big data 
        # geom_point(data = dmt$pts, aes(X, Y), shape = '.', alpha = .1) + 
        # geom_segment(data = dmt$edges[boundary == TRUE], aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt), color = 'red', lwd = .2) + 
    
        ## small data 
        geom_segment(data = dmt$edges, aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt), color = 'black', lwd = .2) + 
        geom_segment(data = dmt$edges[boundary == TRUE], aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt), color = 'red', lwd = .2) + 
        geom_point(data = dmt$pts, aes(X, Y), size = .5) + 
    
        theme_void(base_size = 20) + 
        coord_cartesian(expand = FALSE) + 
        labs(title = 'Pruned adjacency graph') + 
        NULL
}
```

## pca

```{r}
dmt$udv_cells = do_pca(dmt$counts, npcs)
```


# Step 1: compute gradients on all data structures

```{r}
field = compute_gradients(dmt, smooth_distance, smooth_similarity)
field = compress_gradients_svd(field)
```

```{r fig.height = 15, fig.width = 15}
if (show_plots) {    

    len_plot_constant = .8
    fig.size(15, 15)
    # fig.size(20, 20)
    ggplot() + 
        geom_point(data = dmt$pts, aes(X, Y), size = .5) + 
        geom_segment(data = dmt$edges[boundary == TRUE, ], aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt), color = 'red') + 
    
        ## Triangle Gradients
        geom_segment(
            data = data.table(dmt$tris, field$tris_svd), 
            aes(
                x=X-len_plot_constant*(len_grad+len_ortho)*dx_ortho, 
                y=Y-len_plot_constant*(len_grad+len_ortho)*dy_ortho, 
                xend=X+len_plot_constant*(len_grad+len_ortho)*dx_ortho, 
                yend=Y+len_plot_constant*(len_grad+len_ortho)*dy_ortho
            ), 
            linewidth = .4, alpha = 1, 
            color = 'blue'
        ) + 
    
        theme_void() + 
        coord_fixed(expand = FALSE) + 
        NULL
}
```

# Step 2: DMT

## compute f

```{r}
dmt = dmt_set_f(dmt, field)
```

```{r fig.height = 15, fig.width = 15}
if (show_plots) {    
    ntri = max(which(dmt$tris$external == FALSE))
    i = Matrix::t(dmt$tri_to_pt[1:ntri, ])@i+1
    plt_df = data.table(
        X = dmt$pts$X[i],
        Y = dmt$pts$Y[i],
        f = rep(dmt$tris$f[1:ntri], each = 3)
    )[
        , id := rep(1:ntri, each = 3)
    ][]
    
    
    fig.size(15, 15)
    ggplot() + 
        geom_polygon(data = plt_df, aes(X, Y, group = id, fill = f, color = f)) + 
        theme_void() + 
        coord_fixed(expand = FALSE) + 
        scale_fill_viridis() + 
        scale_color_viridis() + 
        NULL
}
```

## forests

```{r}
dmt$prim = do_primary_forest(dmt)
dmt$dual = do_dual_forest(dmt)
```

```{r fig.height = 15, fig.width = 15}
if (show_plots) {    
    fig.size(15, 15)
    ggplot() +     
        ## primary forest
        geom_point(data = dmt$tris[dmt$dual$maxima, ], aes(X, Y), color = 'blue', size = 2) + 
        geom_segment(data = dmt$dual$edges, aes(x=x0, y=y0, xend=x1, yend=y1), color = 'blue') + 

        ## primary forest
        geom_point(data = dmt$pts[dmt$prim$minima, ], aes(X, Y), color = 'red', size = 2) + 
        geom_segment(data = dmt$prim$edges, aes(x=x0, y=y0, xend=x1, yend=y1), color = 'red') + 
    
        theme_void() + 
        coord_cartesian(expand = FALSE) + 
        NULL
}
```

## extract epaths

```{r}
dmt$e_sep = dmt_get_separatrices(dmt)
```

After DMT, we have a separatrix (in blue) and the primary forest connecting points (in red). 

```{r fig.height = 15, fig.width = 15}
if (show_plots) {    
    fig.size(15, 15)
    ggplot() + 
    
        geom_segment(data = dmt$edges[dmt$e_sep, ], aes(x = x0_tri, y = y0_tri, xend = x1_tri, yend = y1_tri), lwd = 1, color = 'blue') + 
        geom_segment(data = dmt$edges[boundary == TRUE], aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt), color = 'blue', lwd = 1) + 
    
        ## primary forest
        geom_point(data = dmt$pts[dmt$prim$minima, ], aes(X, Y), color = 'red', size = 2) + 
        geom_segment(data = dmt$prim$edges, aes(x=x0, y=y0, xend=x1, yend=y1), color = 'red') + 
    
        theme_void() + 
        coord_cartesian(expand = FALSE) + 
        NULL
}
```

## extract tiles


```{r}
dmt = dmt_assign_tiles(dmt)
aggs = dmt_init_tiles(dmt)
```

```{r fig.height = 15, fig.width = 20}
if (show_plots) {    
    set.seed(2)
    fig.size(15, 20)
    ggplot() + 
        geom_sf(data = aggs$meta_data$shape) + 
        # geom_point(data = dmt$pts, aes(X, Y, color = factor(agg_id, sample(nrow(aggs$meta_data)))), size = 1) + 
        # scale_color_tableau() + 
        theme_void() + 
        coord_sf(expand = FALSE) + 
        # coord_cartesian(expand = FALSE) + 
        guides(color = 'none') + 
        
        NULL 
}
```

```{r fig.height = 15, fig.width = 20}
if (show_plots) {    
    set.seed(2)
    fig.size(15, 20)
    ggplot() + 
        geom_sf(data = aggs$meta_data$shape) + 
        geom_point(data = dmt$pts, aes(X, Y, color = type)) + 
        theme_void() + 
        coord_sf(expand = FALSE) + 
        scale_color_tableau() + 
        guides(color = 'none') + 
        NULL 
}
```

# Step 3: Aggregation

## Merge main

First, merge similar aggregates that are nearby. 

```{r}
aggs = init_scores(aggs, agg_mode=2, alpha=alpha, max_npts=max_npts)
aggs = merge_aggs(aggs, agg_mode=2, max_npts=max_npts)
dmt = update_dmt_aggid(dmt, aggs)
aggs = update_agg_shapes(dmt, aggs)
```

## Merge small outliers

Then go ahead and merge small clusters that are smaller than 

```{r}
aggs = init_scores(aggs, agg_mode=3, alpha=alpha, min_npts=min_npts)
aggs = merge_aggs(aggs, agg_mode=3, min_npts=min_npts)
dmt = update_dmt_aggid(dmt, aggs)
aggs = update_agg_shapes(dmt, aggs)
```

## Final tiles 

```{r fig.height = 10, fig.width = 30}
fig.size(10, 30)
if (show_plots) { 
    purrr::map(1:3, function(i) {
        ggplot(cbind(aggs$meta_data, val=aggs$pcs[, i])) + 
            geom_sf(aes(geometry = shape, fill = val)) + 
            theme_void(base_size = 16) + 
            coord_sf(expand = FALSE) + 
            scale_fill_gradient2_tableau() + 
            guides(color = 'none') + 
            labs(title = paste0('PC', i)) + 
            NULL 
    }) %>% 
        purrr::reduce(`|`)
}
```

# Results

## Aggregates 

The primary output is the tiles. Each tile has a row in the meta_data table: 

- npts denotes the number of cells in the tile. 


```{r}
head(aggs$meta_data)
```

We also have pooled gene counts, for differential gene expression analysis. 

```{r}
aggs$counts[1:5, 1:5]
```

And we have PCA embeddings for the tiles. 

```{r}
head(aggs$pcs)
```

The rest of the fields are internal to the algorithm and can be ignored. 

```{r}
setdiff(names(aggs), c('pcs', 'meta_data', 'counts'))
```

## Points

We also keep information about the cells in `dmt`. Most of the fields are duplicates of the inputs and some intermediate results. 


```{r}
names(dmt)
```

The important fields to keep for further analyses are: 

- ORIG_ID: the index of the cell in the input data. Some cells get filtered out as outliers, so not all input cells get assigned to an aggregate. 
- agg_id: the index of the tile in the `aggs` data structures above. 


```{r}
head(dmt$pts[, .(ORIG_ID, agg_id)])
```

# Session Info

```{r}
sessionInfo()
```

