---
title: "Pipeline Walkthrough: Tessera Step-by-Step"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipeline Walkthrough: Tessera Step-by-Step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width='80%',
  fig.align='center'
)
```

# Overview

The Tessera pipeline produces six intermediate data structures, one per stage:

```
make_cells()    →  cells
make_mesh()     →  mesh
compute_field() →  field
compute_morse() →  mesh$morse   (assigned back onto mesh)
run_dmt()       →  dmt_run
make_tiles()    →  tiles_init   (initial DMT tiles)
merge_tiles()   →  tiles        (agglomeratively merged)
```

This vignette runs each stage on a small example dataset, inspects every
slot, and visualises intermediate results so you can see exactly what each
object contains, how they relate to one another, and what the algorithm is
doing at each step.

# Libraries and data

```{r}
suppressPackageStartupMessages({
    library(tessera)

    ## Plotting (not imported by tessera)
    library(ggplot2)
    library(ggthemes)
    library(viridis)
    library(patchwork)
})

## Jupyter plot sizing helper
fig.size <- function(h, w) {
    options(repr.plot.height = h, repr.plot.width = w)
}

## Helper: convert sf tile shapes to geom_polygon-compatible data.table
sf_to_poly <- function(df, shape_col = "shape") {
    shapes <- df[[shape_col]]
    coords_list <- lapply(seq_along(shapes), function(i) {
        co <- sf::st_coordinates(shapes[[i]])
        data.table::data.table(x = co[, 1], y = co[, 2], .tile_group = i)
    })
    coords_dt <- data.table::rbindlist(coords_list)
    meta <- data.table::as.data.table(df)[, !..shape_col]
    meta[, .tile_group := .I]
    merge(coords_dt, meta, by = ".tile_group")
}
```

```{r}
data('tessera_warmup')
counts    = tessera_warmup$counts
meta_data = tessera_warmup$meta_data
cat('Input: ', nrow(meta_data), 'cells,', nrow(counts), 'genes\n')
```

```{r}
## Parameters used throughout
npcs                  = 10
prune_thresh_quantile = 0.95
prune_min_cells       = 10
smooth_distance       = 'projected'
smooth_similarity     = 'projected'
max_npts              = 50
min_npts              = 5
alpha                 = 1
```

```{r fig.height = 8, fig.width = 8}
fig.size(8, 8)
ggplot() +
    geom_point(data = meta_data, aes(X, Y, color = type)) +
    theme_void() +
    scale_color_tableau() +
    coord_fixed() +
    labs(title = 'Input: cell types in physical space') +
    NULL
```

```{r}
table(meta_data$type)
```

---

# 1. `cells` — raw inputs {#cells}

`make_cells()` packages raw inputs into a canonical container that all
downstream stages consume.

```{r}
coords = matrix(
    c(meta_data$X, meta_data$Y), ncol = 2,
    dimnames = list(NULL, c('x', 'y'))
)
cells = make_cells(coords, NULL, counts, data.table::as.data.table(meta_data))
names(cells)
```

| Slot | Type | Description |
|------|------|-------------|
| `coords` | N × 2 matrix | Spatial coordinates (x, y) |
| `embeddings` | named list | `NULL` until PCA is computed after pruning |
| `counts` | G × N dgCMatrix | Raw gene counts |
| `meta` | data.table (N rows) | Per-cell metadata |

```{r}
dim(cells$coords)
head(cells$coords)
```

```{r}
cells$embeddings    # NULL until we call do_pca below
```

```{r}
dim(cells$counts)   # genes x cells
cells$counts[1:4, 1:4]
```

```{r}
dim(cells$meta)
head(cells$meta)
```

---

# 2. `mesh` — Delaunay triangulation {#mesh}

`make_mesh()` triangulates cell coordinates, prunes long edges and small
connected components, and adds degenerate exterior triangles along the
boundary.

```{r}
mesh = make_mesh(cells, list(
    prune_thresh_quantile = prune_thresh_quantile,
    prune_mincells        = prune_min_cells
))
names(mesh)
```

## `mesh$pts` — surviving cells

Data.table of cells that survived pruning — fewer rows than `cells$coords`
when outlier cells are removed.

```{r}
dim(mesh$pts)
names(mesh$pts)
head(mesh$pts)
```

`ORIG_ID` is a 1-based index into the original `cells$coords` and
`cells$meta`:

```{r}
cells$meta$type[mesh$pts$ORIG_ID[1:6]]
```

## `mesh$edges` — primal and dual edges

One row per edge of the triangulation. Every edge has a **primal** form
(connecting two points) and a **dual** form (connecting the centroids of its
two flanking triangles).

```{r}
dim(mesh$edges)
names(mesh$edges)
```

| Column | Description |
|--------|-------------|
| `from_pt`, `to_pt` | Row indices into `mesh$pts` |
| `from_tri`, `to_tri` | Row indices into `mesh$triangles` |
| `x0_pt`, `y0_pt`, `x1_pt`, `y1_pt` | Primal endpoint coordinates |
| `x0_tri`, `y0_tri`, `x1_tri`, `y1_tri` | Dual endpoint coordinates |
| `boundary` | `TRUE` for edges on the convex-hull boundary |

```{r}
table(mesh$edges$boundary)
```

## `mesh$triangles`

One row per triangle, including degenerate exterior triangles.

```{r}
dim(mesh$triangles)
names(mesh$triangles)
table(mesh$triangles$external)
```

| Column | Description |
|--------|-------------|
| `X`, `Y` | Triangle centroid |
| `area` | Triangle area |
| `height` | Circumradius (used for pruning) |
| `external` | `TRUE` for degenerate boundary triangles |

## `mesh$tri_to_pt` and `mesh$adj`

```{r}
dim(mesh$tri_to_pt)   # num_tris x num_pts — sparse incidence matrix
dim(mesh$adj)         # num_pts x num_pts — symmetric adjacency matrix
```

## Visualising the mesh

The pruned adjacency graph (black) with boundary edges highlighted (red).
Points that were pruned as outliers are absent.

```{r fig.height = 15, fig.width = 15}
fig.size(15, 15)
ggplot() +
    geom_segment(data = mesh$edges,
                 aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt),
                 color = 'black', lwd = .2) +
    geom_segment(data = mesh$edges[boundary == TRUE],
                 aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt),
                 color = 'red', lwd = .2) +
    geom_point(data = mesh$pts, aes(X, Y), size = .5) +
    theme_void(base_size = 20) +
    coord_cartesian(expand = FALSE) +
    labs(title = 'Pruned Delaunay mesh (boundary edges in red)') +
    NULL
```

---

# 3. PCA on pruned cells

**Important:** PCA must be computed on the *pruned* cells (those that
survived `make_mesh()`), not the full input.  The resulting embedding matrix
has `nrow(mesh$pts)` rows — one per surviving cell.

```{r}
pca              = do_pca(cells$counts[, mesh$pts$ORIG_ID, drop = FALSE], npcs)
cells$embeddings = list(pca = pca$embeddings)
dim(cells$embeddings$pca)   # num_surviving_cells x npcs
```

---

# 4. `field` — spatial gradient field {#field}

`compute_field()` estimates the transcriptional gradient at every mesh
element (points, triangles, edges) and SVD-compresses each gradient to six
scalars.

```{r}
field = compute_field(cells, mesh, list(
    smooth_distance   = smooth_distance,
    smooth_similarity = smooth_similarity
))
names(field)
names(field$gradient)
```

## SVD gradient columns

Every gradient matrix has the same six columns:

```{r}
colnames(field$gradient$pts)
```

| Column | Description |
|--------|-------------|
| `dx_grad`, `dy_grad` | Principal direction of transcriptional change |
| `dx_ortho`, `dy_ortho` | Direction orthogonal to gradient (isovalue direction) |
| `len_grad` | Gradient magnitude — large near tissue boundaries |
| `len_ortho` | Orthogonal singular value |

## Gradient matrix dimensions

| Slot | Rows | Content |
|------|------|---------|
| `gradient$pts` | num_pts | One gradient per surviving cell |
| `gradient$tris` | num_tris | Weighted average of vertex gradients |
| `gradient$edges_pts` | num_edges | Primal edge gradient |
| `gradient$edges_tris` | num_edges | Dual edge gradient |

```{r}
dim(field$gradient$pts)
dim(field$gradient$tris)
dim(field$gradient$edges_pts)
dim(field$gradient$edges_tris)
```

```{r}
head(field$gradient$pts)
```

## Visualising the gradient field

Triangle gradients visualised as line segments whose length is proportional
to `len_grad + len_ortho`.  Strong gradients (long segments) appear at tissue
boundaries where gene expression changes rapidly.

```{r fig.height = 15, fig.width = 15}
len_plot_constant = .8
fig.size(15, 15)
ggplot() +
    geom_point(data = mesh$pts, aes(X, Y), size = .5) +
    geom_segment(data = mesh$edges[boundary == TRUE, ],
                 aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt),
                 color = 'red') +
    geom_segment(
        data = data.table(mesh$triangles, field$gradient$tris),
        aes(
            x    = X - len_plot_constant * (len_grad + len_ortho) * dx_ortho,
            y    = Y - len_plot_constant * (len_grad + len_ortho) * dy_ortho,
            xend = X + len_plot_constant * (len_grad + len_ortho) * dx_ortho,
            yend = Y + len_plot_constant * (len_grad + len_ortho) * dy_ortho
        ),
        linewidth = .4, alpha = 1,
        color = 'blue'
    ) +
    theme_void() +
    coord_fixed(expand = FALSE) +
    labs(title = 'Gradient field (blue) with boundary edges (red)') +
    NULL
```

---

# 5. `mesh$morse` — Discrete Morse Theory {#morse}

`compute_morse()` derives a scalar field `f` from gradient magnitudes, builds
two spanning forests (primal on points, dual on triangles), and traces the
separatrix edges that define tile boundaries.

The result is stored back on the mesh as `mesh$morse`.

```{r}
mesh$morse = compute_morse(field, mesh)
names(mesh$morse)
```

## `mesh$morse$f` — scalar boundary-score field

`f = len_grad + len_ortho`.  High values indicate tissue boundaries.

```{r}
names(mesh$morse$f)
```

| Slot | Length | Description |
|------|--------|-------------|
| `f$pts` | num_pts | Boundary score at each cell |
| `f$tris` | num_tris | Boundary score at each triangle |
| `f$edges_prim` | num_edges | Primal edge boundary score |
| `f$edges_dual` | num_edges | Dual edge boundary score |

```{r}
length(mesh$morse$f$pts)
summary(mesh$morse$f$pts)
```

Visualising the scalar field as a heatmap on the triangulation.  Bright
(yellow) regions are tissue boundaries; dark (purple) regions are interior:

```{r fig.height = 15, fig.width = 15}
ntri = max(which(mesh$triangles$external == FALSE))
i = Matrix::t(mesh$tri_to_pt[1:ntri, ])@i + 1
plt_df = data.table::data.table(
    X = mesh$pts$X[i],
    Y = mesh$pts$Y[i],
    f = rep(mesh$morse$f$tris[1:ntri], each = 3)
)[
    , id := rep(1:ntri, each = 3)
][]

fig.size(15, 15)
ggplot() +
    geom_polygon(data = plt_df, aes(X, Y, group = id, fill = f, color = f)) +
    theme_void() +
    coord_fixed(expand = FALSE) +
    scale_fill_viridis() +
    scale_color_viridis() +
    labs(title = 'Scalar field f (boundary score) on triangles') +
    NULL
```

## `mesh$morse$prim` and `mesh$morse$dual` — spanning forests

* **Primal forest** (red): minimum spanning forest on points.  Its minima are
  the "seed" cells — one per initial tile.
* **Dual forest** (blue): maximum spanning forest on triangles.  Its maxima
  are the "ridge" triangles sitting at peaks of the boundary score field.

```{r}
names(mesh$morse$prim)
length(mesh$morse$prim$minima)   # number of initial tile seeds
```

```{r}
names(mesh$morse$dual)
length(mesh$morse$dual$maxima)   # number of ridge triangles
```

```{r fig.height = 15, fig.width = 15}
fig.size(15, 15)
ggplot() +
    ## dual forest (blue)
    geom_point(data = mesh$triangles[mesh$morse$dual$maxima, ],
               aes(X, Y), color = 'blue', size = 2) +
    geom_segment(data = mesh$morse$dual$edges,
                 aes(x = x0, y = y0, xend = x1, yend = y1), color = 'blue') +
    ## primal forest (red)
    geom_point(data = mesh$pts[mesh$morse$prim$minima, ],
               aes(X, Y), color = 'red', size = 2) +
    geom_segment(data = mesh$morse$prim$edges,
                 aes(x = x0, y = y0, xend = x1, yend = y1), color = 'red') +
    theme_void() +
    coord_cartesian(expand = FALSE) +
    labs(title = 'Primal forest (red, on points) and dual forest (blue, on triangles)') +
    NULL
```

## `mesh$morse$e_sep` — separatrix edges

Integer vector of 1-based indices into `mesh$edges`.  These edges run along
tissue boundaries and are removed by `run_dmt()` to partition cells into
initial tiles.

```{r}
length(mesh$morse$e_sep)
## Separatrix edges as a fraction of all edges:
length(mesh$morse$e_sep) / nrow(mesh$edges)
```

After DMT, we have the separatrix (blue) cutting through high-f regions, and
the primal forest (red) connecting cells within each tile:

```{r fig.height = 15, fig.width = 15}
fig.size(15, 15)
ggplot() +
    ## separatrix (blue)
    geom_segment(data = mesh$edges[mesh$morse$e_sep, ],
                 aes(x = x0_tri, y = y0_tri, xend = x1_tri, yend = y1_tri),
                 lwd = 1, color = 'blue') +
    geom_segment(data = mesh$edges[boundary == TRUE],
                 aes(x = x0_pt, y = y0_pt, xend = x1_pt, yend = y1_pt),
                 color = 'blue', lwd = 1) +
    ## primal forest (red)
    geom_point(data = mesh$pts[mesh$morse$prim$minima, ],
               aes(X, Y), color = 'red', size = 2) +
    geom_segment(data = mesh$morse$prim$edges,
                 aes(x = x0, y = y0, xend = x1, yend = y1), color = 'red') +
    theme_void() +
    coord_cartesian(expand = FALSE) +
    labs(title = 'Separatrices (blue) and primal forest (red)') +
    NULL
```

---

# 6. `dmt_run` — initial tile assignments {#dmt_run}

`run_dmt()` removes the separatrix edges from the mesh graph and labels each
resulting connected component as a distinct initial tile.

```{r}
dmt_run = run_dmt(mesh)
names(dmt_run)
```

## `dmt_run$pts` — cells with tile IDs

Copy of `mesh$pts` with one added column:

| Column | Description |
|--------|-------------|
| `agg_id` | Integer initial tile ID |

```{r}
names(dmt_run$pts)
head(dmt_run$pts)
```

```{r}
cat('Surviving cells: ', nrow(dmt_run$pts), '\n')
cat('Initial tiles:   ', length(unique(dmt_run$pts$agg_id)), '\n')
```

The initial tiles are fine-grained — many will be merged in `make_tiles()`.

## `dmt_run$edges` — edges with tile labels

Copy of `mesh$edges` with two added columns:

| Column | Description |
|--------|-------------|
| `agg_from` | Tile ID of the `from_pt` endpoint |
| `agg_to` | Tile ID of the `to_pt` endpoint |

```{r}
setdiff(names(dmt_run$edges), names(mesh$edges))
head(dmt_run$edges[, .(from_pt, to_pt, agg_from, agg_to)])
```

## Visualising initial tiles

Each colour is one initial tile.  Note how the separatrices from the previous
step form the boundaries between tiles:

```{r fig.height = 15, fig.width = 20}
set.seed(2)
fig.size(15, 20)
ggplot() +
    geom_point(data = dmt_run$pts, aes(X, Y, color = factor(agg_id)), size = 1) +
    theme_void() +
    coord_cartesian(expand = FALSE) +
    guides(color = 'none') +
    labs(title = 'Initial DMT tiles (one colour per tile)') +
    NULL
```

For comparison, here are the original cell types — the initial tiles should
roughly respect cell-type boundaries:

```{r fig.height = 15, fig.width = 20}
pts_with_meta = data.table::copy(dmt_run$pts)
pts_with_meta[, type := meta_data$type[ORIG_ID]]
set.seed(2)
fig.size(15, 20)
ggplot() +
    geom_point(data = pts_with_meta, aes(X, Y, color = type)) +
    theme_void() +
    coord_cartesian(expand = FALSE) +
    scale_color_tableau() +
    labs(title = 'Cell types (for comparison with initial tiles)') +
    NULL
```

---

# 7. `tiles_init` — initial DMT tiles {#tiles_init}

`make_tiles()` extracts one tile per connected component from the DMT
segmentation.  These are the raw, unmerged tiles — typically many more and
smaller than the final tiles.

```{r}
tiles_init = make_tiles(cells, mesh, dmt_run)
cat('Initial tiles:', nrow(tiles_init$meta_data), '\n')
summary(tiles_init$meta_data$npts)
```

The initial tiles have the same structure as the final tiles (meta_data,
edges, pcs, counts, adj, cell_ids), so you can already inspect or visualise
them before merging.

---

# 8. `tiles` — final merged tiles {#tiles}

`merge_tiles()` agglomeratively merges the initial tiles into final spatial
tiles, first merging tiles larger than `max_npts` and then absorbing tiles
smaller than `min_npts` into their most similar neighbour.

```{r}
tiles = merge_tiles(tiles_init, list(
    alpha    = alpha,
    max_npts = max_npts,
    min_npts = min_npts
))
names(tiles)
```

## `tiles$meta_data` — one row per tile

```{r}
names(tiles$meta_data)
head(dplyr::select(tiles$meta_data, -shape))
```

| Column | Description |
|--------|-------------|
| `id` | Tile identifier |
| `X`, `Y` | Tile centroid |
| `npts` | Number of constituent cells |
| `shape` | `sf` MULTIPOLYGON tile boundary |
| `area`, `perimeter` | Tile geometry summaries |

```{r}
cat('Final tiles:', nrow(tiles$meta_data), '\n')
summary(tiles$meta_data$npts)
```

`shape` is an `sf` geometry column:

```{r}
class(tiles$meta_data$shape)
tiles$meta_data$shape[1]
```

## `tiles$counts` — pooled gene counts

Genes × num_tiles sparse matrix.  Column names match `tiles$meta_data$id`.

```{r}
dim(tiles$counts)
identical(colnames(tiles$counts), tiles$meta_data$id)
tiles$counts[1:4, 1:4]
```

## `tiles$pcs` — per-tile PCA embeddings

num_tiles × D matrix.  Each row is the mean of the constituent cells'
PC embeddings.  Row names match `tiles$meta_data$id`.

```{r}
dim(tiles$pcs)
head(tiles$pcs[, 1:4])
```

## `tiles$adj` — tile adjacency matrix

Symmetric dgCMatrix (num_tiles × num_tiles).  Entry (i, j) = 1 if tiles i
and j share a border.

```{r}
dim(tiles$adj)
mean(Matrix::colSums(tiles$adj > 0))   # average number of tile neighbours
```

## `tiles$edges` — inter-tile edges

Data.table with one row per pair of adjacent tiles.

```{r}
dim(tiles$edges)
names(tiles$edges)
head(dplyr::select(tiles$edges, from, to, npts, edge_length))
```

## `tiles$cell_ids` — cell-to-tile mapping

The primary way to transfer tile labels back to individual cells.

```{r}
head(tiles$cell_ids)
cat('Cells assigned to tiles:', nrow(tiles$cell_ids), 'of', nrow(meta_data), '\n')
```

| Column | Description |
|--------|-------------|
| `ORIG_ID` | 1-based index into original `cells$coords` / `meta_data` |
| `tile_id` | Tile identifier (matches `tiles$meta_data$id`) |

## Visualising final tiles

The first three tile PCs capture the dominant spatial patterns of gene
expression variation:

```{r fig.height = 10, fig.width = 30}
fig.size(10, 30)
purrr::map(1:3, function(i) {
    ggplot(sf_to_poly(cbind(tiles$meta_data, val = tiles$pcs[, i]))) +
        geom_polygon(aes(x, y, group = .tile_group, fill = val)) +
        theme_void(base_size = 16) +
        coord_fixed() +
        scale_fill_gradient2_tableau() +
        guides(color = 'none') +
        labs(title = paste0('PC', i)) +
        NULL
}) %>%
    purrr::reduce(`|`)
```

---

# Cross-referencing objects {#cross-ref}

## Transfer tile labels to original cells

```{r}
meta_data$tile_id = NA
meta_data$tile_id[tiles$cell_ids$ORIG_ID] = tiles$cell_ids$tile_id
table(is.na(meta_data$tile_id))   # FALSE = assigned, TRUE = pruned as outlier
```

## All cells in a given tile

```{r}
first_tile = tiles$meta_data$id[1]
cells_in_tile1 = tiles$cell_ids[tile_id == first_tile, ORIG_ID]
meta_data[cells_in_tile1, ]
```

## Cells that survived mesh pruning

```{r}
## mesh$pts$ORIG_ID indexes into the original meta_data
surviving_meta = meta_data[mesh$pts$ORIG_ID, ]
cat('Surviving cells:', nrow(surviving_meta), 'of', nrow(meta_data), '\n')
```

## Object dimension summary

```{r}
cat(sprintf(
    'Input cells:          %d\nSurviving (pruned):   %d\nInitial tiles (DMT):  %d\nFinal tiles (merged): %d\n',
    nrow(meta_data),
    nrow(mesh$pts),
    nrow(tiles_init$meta_data),
    nrow(tiles$meta_data)
))
```

---

# Session Info

```{r}
sessionInfo()
```
